---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
The data has been downloaded from [the course website](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) at 
May 5th 2021 and added to GitHub repository. All the rows with NA in *steps*
column have been removed.

```{r loading}
library(dplyr)
activity_raw <- read.csv("data/activity.csv")
activity <- activity_raw %>%
    filter(!is.na(steps))
head(activity)
```

## What is the mean total number of steps taken per day?

1. Calculate the total number of steps each day
```{r}
total_steps <- activity %>%
    group_by(date) %>%
    summarise(steps_per_day = sum(steps))

```
2. Total number of steps take each day. 
```{r hist}
library(ggplot2)

total_steps %>%
    ggplot(aes(steps_per_day)) + 
    geom_histogram(binwidth = 1000)

```

3. Calculate and report the mean and median of the total number of steps taken 
per day.
```{r}
mean(total_steps$steps_per_day)
```
```{r}
median(total_steps$steps_per_day)
```
## What is the average daily activity pattern?

1. Time series plot of the average number of steps taken

```{r timeseries}
mean_isteps <- activity %>%
     group_by(interval) %>%
     summarise(mean_steps = mean(steps))

 ggplot(mean_isteps, aes(interval, mean_steps)) + geom_line(color = "navy")

```
2. The 5-minute interval that, on average, contains the maximum number of steps.
```{r maxinterval}
mean_isteps$interval[which.max(mean_isteps$mean_steps)]
```

## Imputing missing values
1. Calculate and report the total number of missing values in the dataset.
```{r}
sum(is.na(activity_raw))
```
2. For the missing values, let's put the mean for the same interval. Drop additional
columns created during the process.
```{r}
activity_mod <- activity_raw %>%
    group_by(interval) %>% 
    mutate(mean_steps = mean(steps, na.rm = TRUE)) %>%
    mutate(steps = ifelse(is.na(steps), mean_steps, steps)) %>%
    select(-mean_steps)

```
3. Histogram of the total number of steps taken each day after missing values are imputed.
```{r}
total_mod_steps <- activity_mod %>%
    group_by(date) %>%
    summarise(steps_per_day = sum(steps))

total_mod_steps %>%
    ggplot(aes(steps_per_day)) + 
    geom_histogram(binwidth = 1000)
```

## Are there differences in activity patterns between weekdays and weekends?
1. Calculate if there are weekends. We assume that week starts on Monday. 
```{r}
library(lubridate)
activity_wday <- activity_mod %>%
    mutate(wday = wday(ymd(date), week_start = 1)) %>%
    mutate(is_weekend = ifelse(wday > 5, TRUE, FALSE)) %>%
    group_by(is_weekend, interval) %>%
    summarise(mean_per_wday = mean(steps), .groups = 'drop')
```
2. Let's see how working days (FALSE) are different from weekends (TRUE). 
```{r}
ggplot(activity_wday, aes(interval, mean_per_wday)) +
    geom_line(color = "navy") + 
    facet_grid(rows = vars(is_weekend))
```
